---
title: "MATH252 Group Project"
format:
  pdf:
    keep-tex: true
editor: visual
execute:
  echo: false
  warning: false
---

## Install Libraries and Functions

```{r}
# install libraries
library(tidyverse)
library(factoextra)   # fviz_dist(), hopkins()
library(cluster)      # daisy() for Gower distance on mixed data
library(knitr)
library(kableExtra)
library(hopkins)

library(conflicted)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::mutate)
```

```{r}
# ── Split into continuous and categorical subsets ─────────────────────────────
split_vars <- function(df) {

  df_cont <- df |>
    select(
      starts_with("work_time_in_seconds"),
      multiplier,
      amount,
      starts_with("q3_1"),
      starts_with("q8_1")
    ) |>
    drop_na() |>
    mutate(across(everything(), as.numeric))

  df_cat <- df |>
    select(
      age, gender, income,
      starts_with("q5"),
      starts_with("q9"),
      starts_with("q7"),
      starts_with("q8_aside")
    ) |>
    drop_na() |>
    mutate(across(everything(), as.factor))

  list(continuous = df_cont, categorical = df_cat)
}

# ── Hopkins statistic for both subsets ───────────────────────────────────────
compute_hopkins <- function(split, seed = 420) {
  cont_scaled <- split$continuous |> scale() |> as.data.frame()
  set.seed(seed)
  h_cont <- hopkins(cont_scaled, nrow(cont_scaled) - 1)

  gower_d <- daisy(split$categorical, metric = "gower")
  k_dims  <- min(5, ncol(split$categorical) - 1)
  mds_mat <- cmdscale(gower_d, k = k_dims) |> as.data.frame()
  set.seed(seed)
  h_cat <- hopkins(mds_mat, nrow(mds_mat) - 1)

  list(H_continuous = round(h_cont, 4),
       H_categorical = round(h_cat, 4))
}

# ── VAT plots via fviz_dist() ─────────────────────────────────────────────────
plot_vat <- function(split, label, hop_vals, max_n = 300, seed = 420) {
  cont_data <- split$continuous |> scale() |> as.data.frame()
  if (nrow(cont_data) > max_n) {
    set.seed(seed); cont_data <- cont_data[sample(nrow(cont_data), max_n), ]
  }

  p_cont <- fviz_dist(
    dist(cont_data, method = "euclidean"),
    order = TRUE, show_labels = FALSE,
    gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07")
  ) + labs(title    = paste0("VAT \u2014 Continuous | ", label),
           subtitle = paste("Hopkins H =", hop_vals$H_continuous))

  cat_data <- split$categorical
  if (nrow(cat_data) > max_n) {
    set.seed(seed); cat_data <- cat_data[sample(nrow(cat_data), max_n), ]
  }

  p_cat <- fviz_dist(
    daisy(cat_data, metric = "gower"),
    order = TRUE, show_labels = FALSE,
    gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07")
  ) + labs(title    = paste0("VAT \u2014 Categorical | ", label),
           subtitle = paste("Hopkins H =", hop_vals$H_categorical))

  list(vat_continuous = p_cont, vat_categorical = p_cat)
}

# Create plots folder on load
if (!dir.exists("plots")) dir.create("plots", recursive = TRUE)

# Save a ggplot object as PNG
save_plot_gg <- function(plot_obj, filename, path = "plots/",
                         width = 10, height = 6, dpi = 300) {
  filepath <- file.path(path, paste0(filename, ".png"))
  ggsave(filepath, plot  = plot_obj,
                   width = width,
                   height = height,
                   dpi    = dpi,
                   bg     = "white")
  cat("Saved:", filepath, "\n")
  invisible(filepath)
}

# Save a base R plot expression as PNG
save_plot_base <- function(plot_expr, filename, path = "plots/",
                           width = 3000, height = 1800, res = 300) {
  filepath <- file.path(path, paste0(filename, ".png"))
  png(filepath, width = width, height = height, res = res)
  par(mar = c(5, 4, 4, 10))
  force(plot_expr)
  dev.off()
  cat("Saved:", filepath, "\n")
  invisible(filepath)
}

# ── Master wrapper ────────────────────────────────────────────────────────────
analyse_dataset <- function(df, label) {
  message("Processing: ", label)

  clean_label <- gsub("[^A-Za-z0-9]", "_", label)

  split <- split_vars(df)
  hops  <- compute_hopkins(split)

  # ── Continuous VAT ─────────────────────────────────────────────────────────
  cont_data <- split$continuous |> scale() |> as.data.frame()
  if (nrow(cont_data) > 300) {
    set.seed(123); cont_data <- cont_data[sample(nrow(cont_data), 300), ]
  }

  p_cont <- fviz_dist(
    dist(cont_data, method = "euclidean"),
    order = TRUE, show_labels = FALSE,
    gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07")
  ) + labs(title    = paste0("VAT \u2014 Continuous | ", label),
           subtitle = paste("Hopkins H =", hops$H_continuous))

  # ── Categorical VAT ────────────────────────────────────────────────────────
  cat_data <- split$categorical
  if (nrow(cat_data) > 300) {
    set.seed(123); cat_data <- cat_data[sample(nrow(cat_data), 300), ]
  }

  p_cat <- fviz_dist(
    daisy(cat_data, metric = "gower"),
    order = TRUE, show_labels = FALSE,
    gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07")
  ) + labs(title    = paste0("VAT \u2014 Categorical | ", label),
           subtitle = paste("Hopkins H =", hops$H_categorical))

  # Print to document
  print(p_cont)
  print(p_cat)

  # Save as PNG
  save_plot_gg(p_cont, paste0("vat_continuous_", clean_label))
  save_plot_gg(p_cat,  paste0("vat_categorical_", clean_label))

  list(label = label, hopkins = hops,
       plots  = list(vat_continuous = p_cont,
                     vat_categorical = p_cat))
}
```

## EDA

### Setup & Load Original Dataset

```{r}
path0 <- "~/Downloads/MATH 252/Group Project/Raw Data.csv"

raw <- read.csv(path0, na.strings = c("", "NA", "#N/A"))

# Derive price from the paper's definition: price = 1 / multiplier
# (cost to the dictator of having $1 received by the passive player)
# Rename using dplyr::rename() with the dotted names read.csv() produces
raw <- raw |>
  rename(
    work_time_main     = WorkTimeInSeconds,
    work_time_followup = WorkTimeInSeconds.1,
    political_views    = Q3_1.On.a.scale.of.0.to.100..how.would.you.describe.your.political.views.,
    religious_views    = Q8_1.On.a.scale.of.0.to.100..how.would.you.describe.your.religious.orientation.,
    vote_2016          = Q5.In.the.2016.Presidential.election..who.did.you.vote.for.,
    state              = Q9.What.State.do.you.live.in.,
    party              = Q7.Do.you.consider.yourself.a.,
    attendance         = Q8.Aside.from.weddings.and.funerals..how.often.do.you.attend.religious.services.
  )

# Derive price and gave, retype factors
raw <- raw |>
  mutate(
    price = ifelse(multiplier > 0, 1 / multiplier, NA),
    gave  = as.integer(amount > 0),
    age = factor(age,
      levels  = c("18-29","30-39","40-49","50-59","60-69","70 or over"),
      ordered = TRUE),
    income = factor(income,
      levels  = c("Under $20,000","$20,000 - $39,999","$40,000 - $59,999",
                  "$60,000 - $79,999","$80,000 - $99,999","Over $100,000"),
      ordered = TRUE),
    gender     = factor(gender),
    party      = factor(party),
    vote_2016  = factor(vote_2016),
    attendance = factor(attendance,
      levels  = c("Never","A few times a year","Once or twice a month",
                  "Once a week","More than once a week"),
      ordered = TRUE)
  )
```

### Variable Codebook Table

The dataset merges two MTurk HITs. The **main experiment** (Columns 1–15) collected the dictator game decision and basic demographics from 772 subjects. The **follow-up survey** (Columns 16 onward) collected extended political and religious measures from the 505 subjects who responded — indicated by `#N/A` missingness for non-respondents (Lillehaugen et al., 2023, §2.1)

```{r}
codebook <- tibble(
  Variable = c(
    "WorkTimeInSeconds",
    "multiplier",
    "amount",
    "price",
    "age",
    "gender",
    "income",
    "WorkTimeInSeconds.1",
    "Q3_1...",
    "Q8_1...",
    "Q5...",
    "Q9...",
    "Q7...",
    "Q8.Aside..."
  ),
  Type = c(
    "Continuous","Continuous","Continuous","Continuous",
    "Ordinal","Nominal","Ordinal",
    "Continuous","Continuous","Continuous",
    "Nominal","Nominal","Nominal","Ordinal"
  ),
  Description = c(
    "Seconds to complete the main HIT (data quality indicator)",
    "Multiplication factor μ ~ Uniform(0,2); shown as 0–200%. Scales the amount received by the passive player",
    "Dollars sent to passive player [0, 1]; the primary outcome of the dictator game",
    "Effective price = 1/μ; cost to dictator of the passive player receiving $1 (derived)",
    "Age bracket of subject (6 ordered levels: 18–29 through 70+)",
    "Self-reported gender (Male / Female)",
    "Annual household income bracket (6 ordered levels: Under $20k through Over $100k)",
    "Seconds to complete the follow-up Qualtrics survey (NA = non-respondent)",
    "Political views: 0 = Very liberal, 100 = Very conservative (follow-up survey)",
    "Religious orientation: 0 = Not at all religious, 100 = Very religious (follow-up survey)",
    "Candidate voted for in the 2016 U.S. presidential election (follow-up survey)",
    "U.S. state of residence (follow-up survey)",
    "Political party self-identification (follow-up survey)",
    "Religious service attendance frequency, excluding weddings/funerals (follow-up survey)"
  )
)

kable(codebook, longtable = FALSE,
      caption   = "Variable Codebook — Raw MTurk Dataset",
      col.names = c("Variable", "Type", "Description"),
      booktabs  = TRUE) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down"),
                font_size = 9) |>
  column_spec(3, width = "9cm")
```

### Sample Attrition Table

Of 922 subjects who accepted the HIT, 772 provided valid main-experiment responses (dropout rate 16.3%). Of those, 505 completed the follow-up survey (response rate 65.4%) (Lillehaugen et al., 2023, §2.2.1).

```{r}
n_total      <- nrow(raw)
n_gave_main  <- sum(!is.na(raw$amount))
n_followup   <- sum(!is.na(raw$q3_1))  # proxy: responded to follow-up

attrition <- tibble(
  Stage      = c("Accepted HIT (raw)", "Valid main experiment", "Completed follow-up survey"),
  N          = c(n_total, n_gave_main, n_followup),
  Pct_prev   = c(NA,
                 round(n_gave_main / n_total * 100, 1),
                 round(n_followup  / n_gave_main * 100, 1))
)

kable(attrition, longtable = FALSE,
      caption   = "Sample Attrition Across Experiment Stages",
      col.names = c("Stage", "N", "% of Previous Stage"),
      digits    = 1,
      booktabs  = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Continuous Variable Summary Table

The multiplier is the central experimental manipulation. The `amount` variable is the dictator game outcome. `price` is derived as 1/μ per the paper's elasticity models (Eq. 2, §2.2.5)

```{r}
cont_summary <- raw |>
  dplyr::select(work_time_main, multiplier, amount, price,
                political_views, religious_views, work_time_followup) |>
  pivot_longer(everything(), names_to = "Variable") |>
  group_by(Variable) |>
  summarise(
    N      = sum(!is.na(value)),
    Mean   = round(mean(value, na.rm = TRUE), 3),
    SD     = round(sd(value,   na.rm = TRUE), 3),
    Min    = round(min(value,  na.rm = TRUE), 3),
    Q1     = round(quantile(value, 0.25, na.rm = TRUE), 3),
    Median = round(median(value,   na.rm = TRUE), 3),
    Q3     = round(quantile(value, 0.75, na.rm = TRUE), 3),
    Max    = round(max(value,  na.rm = TRUE), 3),
    `NA`   = sum(is.na(value))
  )

kable(cont_summary, longtable = FALSE,
      caption  = "Descriptive Statistics — Continuous Variables",
      booktabs = TRUE) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down"),
                font_size = 9)
```

### Categorical Frequency Table

```{r}
freq_table <- function(df, var, caption_text) {
  df |>
    filter(!is.na(.data[[var]])) |>
    count(.data[[var]], name = "N") |>
    mutate(Percent = round(N / sum(N) * 100, 1)) |>
    rename(Category = 1) |>
    kable(longtable  = FALSE,
          caption    = caption_text,
          col.names  = c("Category", "N", "%"),
          booktabs   = TRUE) |>
    kable_styling(latex_options = "HOLD_position")
}

freq_table(raw, "gender",     "Gender Distribution")
freq_table(raw, "age",        "Age Distribution")
freq_table(raw, "income",     "Income Distribution")
freq_table(raw, "vote_2016",  "2016 Presidential Vote (Follow-up Respondents)")
freq_table(raw, "party",      "Political Party Affiliation (Follow-up Respondents)")
freq_table(raw, "attendance", "Religious Service Attendance (Follow-up Respondents)")
```

### Dictator Game Output Summary

The paper reports that 39.4% of subjects sent \$0, 15.6% sent the full \$1.00, and the mean amount conditional on giving was \$0.55 (§2.2.3). This block reproduces those figures from the raw data.

```{r}
giving_summary <- raw |>
  filter(!is.na(amount)) |>
  summarise(
    N                  = n(),
    `Mean Amount Sent` = round(mean(amount), 3),
    `SD`               = round(sd(amount), 3),
    `Sent $0 (N)`      = sum(amount == 0),
    `Sent $0 (%)`      = round(mean(amount == 0) * 100, 1),
    `Sent $1.00 (N)`   = sum(amount == 1),
    `Sent $1.00 (%)`   = round(mean(amount == 1) * 100, 1),
    `Mean Given Gave > 0`  = round(mean(amount[amount > 0]), 3)
  ) |>
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value")

kable(giving_summary, longtable = FALSE,
      caption  = "Dictator Game Giving Behavior Summary",
      booktabs = TRUE,
      escape   = FALSE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Multiplier(Price) Distribution

Because μ is drawn from Uniform(0, 2), the price 1/μ is right-skewed. This block checks whether the experimental randomization produced a roughly uniform multiplier distribution.

```{r}
# Binned multiplier distribution
multiplier_dist <- raw |>
  filter(!is.na(multiplier)) |>
  mutate(multiplier_bin = cut(multiplier,
    breaks = seq(0, 2, by = 0.25),
    include.lowest = TRUE, right = FALSE)) |>
  count(multiplier_bin, name = "N") |>
  mutate(Percent = round(N / sum(N) * 100, 1))

kable(multiplier_dist, longtable = FALSE,
      caption   = "Distribution of Multiplier (\u03bc) — Binned in 0.25 Intervals",
      col.names = c("Multiplier Range", "N", "%"),
      booktabs  = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

### Mean Amount Sent by Demographic Group

Table 1 in the paper reports correlations between amount sent and demographic group. This block produces the underlying means for replication.Table 1 in the paper reports correlations between amount sent and demographic group. This block produces the underlying means for replication.

```{r}
# Function: mean amount sent ± SD by a grouping variable
group_giving <- function(df, group_var, label) {
  df |>
    filter(!is.na(amount), !is.na(.data[[group_var]])) |>
    group_by(Group = as.character(.data[[group_var]])) |>  # coerce to character
    summarise(
      N         = n(),
      Mean_Sent = round(mean(amount), 3),
      SD        = round(sd(amount), 3),
      Pct_Gave  = round(mean(gave) * 100, 1)
    ) |>
    mutate(Variable = label) |>
    relocate(Variable)
}

bind_rows(
  group_giving(raw, "gender",    "Gender"),
  group_giving(raw, "age",       "Age"),
  group_giving(raw, "income",    "Income"),
  group_giving(raw, "party",     "Party"),
  group_giving(raw, "vote_2016", "2016 Vote")
) |>
  kable(longtable = FALSE,
        caption   = "Mean Amount Sent and Probability of Giving by Demographic Group",
        col.names = c("Variable", "Group", "N", "Mean Sent ($)", "SD", "% Gave"),
        booktabs  = TRUE) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down"),
                font_size = 9) |>
  collapse_rows(columns = 1, latex_hline = "major")
```

### Missing Data Summary

```{r}
missing_summary <- raw |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(),
               names_to  = "Variable",
               values_to = "N_Missing") |>
  mutate(Pct_Missing = round(N_Missing / nrow(raw) * 100, 1)) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))

kable(missing_summary, longtable = FALSE,
      caption   = "Missing Data by Variable",
      col.names = c("Variable", "N Missing", "% Missing"),
      booktabs  = TRUE) |>
  kable_styling(latex_options = "HOLD_position")
```

## Check Cluster Tendency

### Complete Data

```{r}
path1   <- "~/Downloads/MATH 252/Group Project/processed data/data_complete_baseline.csv"
complete <- read.csv(path1)

res_complete <- analyse_dataset(complete, label = "Complete Data")
```

### Complete Baseline

```{r}
path2        <- "~/Downloads/MATH 252/Group Project/processed data/data_complete.csv"
complete_base <- read.csv(path2)

res_complete_base <- analyse_dataset(complete_base, label = "Complete Baseline")
```

### Mean Mode Imputed

```{r}
path3    <- "~/Downloads/MATH 252/Group Project/processed data/mean_mode_imputed_dataset.csv"
mean_mode <- read.csv(path3)

res_mean_mode <- analyse_dataset(mean_mode, label = "Mean/Mode Imputed")
```

### MICE Imputed 1

```{r}
path4  <- "~/Downloads/MATH 252/Group Project/processed data/mice_imputation_data1.csv"
mice1  <- read.csv(path4)

res_mice1 <- analyse_dataset(mice1, label = "MICE Imputed 1")
```

### MICE Imputed 2

```{r}
path5  <- "~/Downloads/MATH 252/Group Project/processed data/mice_imputation_data2.csv"
mice2  <- read.csv(path5)

res_mice2 <- analyse_dataset(mice2, label = "MICE Imputed 2")
```

### KNN Imputed (K=5)

```{r}
path6  <- "~/Downloads/MATH 252/Group Project/processed data/knn_imputed_data_k5.csv"
knn_k5 <- read.csv(path6)

res_knn_k5 <- analyse_dataset(knn_k5, label = "KNN Imputed (k=5)")
```

### KNN Imputed (K=10)

```{r}
path7   <- "~/Downloads/MATH 252/Group Project/processed data/knn_imputed_data_k10.csv"
knn_k10 <- read.csv(path7)

res_knn_k10 <- analyse_dataset(knn_k10, label = "KNN Imputed (k=10)")
```

### Cluster Tendency Summary

```{r}
hopkins_summary <- map_dfr(
  list(res_complete, res_complete_base, res_mean_mode,
       res_mice1, res_mice2, res_knn_k5, res_knn_k10),
  \(r) tibble(
    Dataset       = r$label,
    H_Continuous  = r$hopkins$H_continuous,
    H_Categorical = r$hopkins$H_categorical
  )
)

knitr::kable(
  hopkins_summary,
  caption     = "Hopkins Statistic by Dataset and Variable Type",
  col.names   = c("Dataset", "H (Continuous)", "H (Categorical)"),
  digits      = 4
)
```

## Analysis Time!

come along with me and the butterflies and bees we can wander through the forest and do so as we please

### Additional libraries

```{r}
library(fclust)         # FKM()
library(FPDclustering)  # PDC()
library(MixGHD)         # ARI()
library(clValid)        # Dunn index
library(factoextra)     # fviz_nbclust()
library(cluster)        # silhouette(), agnes(), daisy()
library(flextable)
library(magick)
```

### Data Preparation

```{r}
# Computes pairwise ARI between datasets within each method
# all_clust_labels: named list of label lists, one per dataset
cross_dataset_ari <- function(all_clust_labels) {

  methods      <- c("km", "fkm", "pdc", "hc")
  method_names <- c("K-Means", "Fuzzy K-Means",
                    "PD Clustering", "Hierarchical Avg")
  dataset_names <- names(all_clust_labels)
  n_datasets    <- length(dataset_names)

  # All pairwise dataset combinations
  pairs <- combn(dataset_names, 2, simplify = FALSE)

  results <- map_dfr(seq_along(methods), function(m_idx) {
    method     <- methods[m_idx]
    method_name <- method_names[m_idx]

    map_dfr(pairs, function(pair) {

      d1 <- pair[1]; d2 <- pair[2]
      labs1 <- all_clust_labels[[d1]][[method]]
      labs2 <- all_clust_labels[[d2]][[method]]

      # Align lengths — take minimum if datasets differ in size
      min_n <- min(length(labs1), length(labs2))
      ari_val <- tryCatch(
        ARI(labs1[1:min_n], labs2[1:min_n]),
        error = function(e) NA_real_
      )

      tibble(
        Method     = method_name,
        Dataset_1  = d1,
        Dataset_2  = d2,
        ARI        = round(ari_val, 4)
      )
    })
  })

  results
}

# Save a kable table as a PNG file
# Returns the filepath for reference
save_table_png <- function(data, filename, caption_text, 
                           path = "tables/",
                           group_col = NULL,    # column name to group by
                           group_size = NULL) { # rows per group

  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
  filepath <- file.path(path, paste0(filename, ".png"))

  df <- as.data.frame(data)

  ft <- flextable::flextable(df) |>
    flextable::set_caption(caption_text) |>
    flextable::theme_booktabs() |>
    flextable::fontsize(size = 11, part = "all") |>
    flextable::font(fontname = "Times New Roman", part = "all") |>
    flextable::bold(part = "header") |>
    flextable::autofit()

  # ── Add dataset group separators if requested ─────────────────────────────
  if (!is.null(group_col) && !is.null(group_size)) {

    n_groups    <- length(unique(df[[group_col]]))
    group_names <- unique(df[[group_col]])
    n_rows      <- nrow(df)

    # Add thick border line between each dataset group
    for (i in 1:(n_groups - 1)) {
      sep_row <- i * group_size
      ft <- ft |>
        flextable::hline(i         = sep_row,
                         border    = officer::fp_border(width = 1.5),
                         part      = "body")
    }

    # Bold the first row of each group and add group label in background
    for (i in 1:n_groups) {
      first_row <- (i - 1) * group_size + 1
      ft <- ft |>
        flextable::bold(i = first_row, part = "body") |>
        flextable::bg(i   = seq(first_row,
                                first_row + group_size - 1),
                      bg  = ifelse(i %% 2 == 0,
                                   "#F5F5F5", "#FFFFFF"),  # alternating shading per group
                      part = "body")
    }

    # Hide the group column from display (already encoded in grouping)
    ft <- ft |>
      flextable::bold(j = group_col, part = "body") |>
      flextable::color(j     = group_col,
                       color = "#2C3E50",
                       part  = "body")
  }

  flextable::save_as_image(ft, path = filepath, res = 300)
  cat("Saved:", filepath, "\n")
  invisible(filepath)
}

# Extract and scale continuous variables for clustering
# Mirrors the split_vars() logic already defined above
prep_cluster_data <- function(df) {
  df |>
    dplyr::select(
      starts_with("work_time_in_seconds"),
      multiplier,
      amount,
      starts_with("q3_1"),
      starts_with("q8_1")
    ) |>
    drop_na() |>
    mutate(across(everything(), as.numeric)) |>
    scale() |>
    as.data.frame()
}

dataset_list <- list(
  list(df = complete,       label = "Complete Data"),
  list(df = complete_base,  label = "Complete Baseline"),
  list(df = mean_mode,      label = "Mean/Mode Imputed"),
  list(df = mice1,          label = "MICE Imputed 1"),
  list(df = mice2,          label = "MICE Imputed 2"),
  list(df = knn_k5,         label = "KNN k=5"),
  list(df = knn_k10,        label = "KNN k=10")
)
```

### CH Index Function

Taken directly from Dr. Tortora's lecture (kmeans.r, lines 4–18).

```{r}
# Calinski-Harabasz ratio — from lecture kmeans.r
ratio.strength <- function(X, G, output) {
  n   <- nrow(X)
  sst <- cov.wt(X)$cov * (n - 1)
  ssw <- 0
  for (g in 1:G) {
    mu.g      <- output$centers[g, ]
    partition <- which(g == output$cluster)
    ng        <- length(partition)
    ssw       <- ssw + cov.wt(X[partition, ], center = mu.g)$cov * (ng - 1)
  }
  ssb <- sst - ssw
  val <- (sum(diag(ssb)) * (n - G)) / (sum(diag(ssw)) * (G - 1))
  return(val)
}
```

### Optimal K: Gap Statistic and Silhouette

Follows `fviz_nbclust()` usage from lecture (clustervalidation.R, lines 33–45). The `FKM2` wrapper is taken directly from Dr. Tortora's lecture (clustervalidation.R, lines 39–43).

```{r}
#| fig-height: 5
#| fig-width: 8
#| fig-keep: all

FKM2 <- function(data, k) {
  res    <- FKM(data, k)
  output <- list(cluster = as.vector(res$clus[, 1]))
  return(output)
}

# Extract optimal K from average silhouette width
# Returns the k in 2:k_max with highest mean silhouette
optimal_k_silhouette <- function(data, FUN, k_max = 8, seed = 420, m = 2) {
  set.seed(seed)
  dist_mat  <- dist(data, method = "euclidean")
  sil_scores <- rep(NA, k_max)

  for (k in 2:k_max) {
    set.seed(seed)
    clust <- FUN(data, k)
    sil_scores[k] <- mean(silhouette(clust$cluster, dist_mat)[, 3])
  }
  which.max(sil_scores)
}

# Extract optimal K from gap statistic
# Returns the k selected by firstSEmax criterion
optimal_k_gap <- function(data, FUN, k_max = 8, seed = 420) {
  set.seed(seed)
  gap_res <- clusGap(data, FUNcluster = FUN,
                     K.max = k_max, B = 50, verbose = FALSE)
  maxSE(gap_res$Tab[, "gap"], gap_res$Tab[, "SE.sim"],
        method = "firstSEmax")
}

# PD Clustering wrapper for clusGap and silhouette
PDC2 <- function(data, k) {
  res    <- PDC(data, k = k)
  output <- list(cluster = as.integer(res$label))
  return(output)
}

# Hierarchical average wrapper
HC2 <- function(data, k) {
  hc  <- agnes(data, method = "average")
  cut <- cutree(as.hclust(hc), k = k)
  list(cluster = as.integer(cut))
}

# Get optimal K for all four methods on one dataset
# Uses silhouette as primary criterion (interpretable across all methods)
get_optimal_k <- function(data, label, k_max = 8, seed = 420) {
  
  clean_label <- gsub("[^A-Za-z0-9]", "_", label)
  set.seed(seed)

  cat("  Computing optimal K — K-Means...\n")
  k_km  <- optimal_k_silhouette(data,
              FUN = function(d, k) {
                set.seed(seed)
                list(cluster = kmeans(d, centers = k, nstart = 50)$cluster)
              },
              k_max = k_max, seed = seed)

  cat("  Computing optimal K — Fuzzy K-Means...\n")
  k_fkm <- optimal_k_silhouette(data,
              FUN = function(d, k) {
                set.seed(seed)
                list(cluster = as.integer(FKM(d, k = k, m = 2)$clus[, 1]))
              },
              k_max = k_max, seed = seed)

  cat("  Computing optimal K — PD Clustering...\n")
  k_pdc <- optimal_k_silhouette(data,
              FUN = function(d, k) PDC2(d, k),
              k_max = k_max, seed = seed)

  cat("  Computing optimal K — Hierarchical Avg...\n")
  k_hc  <- optimal_k_silhouette(data,
              FUN = function(d, k) HC2(d, k),
              k_max = k_max, seed = seed)

  # ── Gap statistic plots ───────────────────────────────────────────────────
  set.seed(seed)
  p_gap_km <- fviz_nbclust(data, kmeans, method = "gap_stat",
                            k.max = k_max, nboot = 50) +
    labs(title = paste("Gap Statistic — K-Means |", label))

  set.seed(seed)
  p_gap_fkm <- fviz_nbclust(data, FKM2, method = "gap_stat",
                             k.max = k_max, nboot = 50) +
    labs(title = paste("Gap Statistic — Fuzzy K-Means |", label))

  # ── Average silhouette width plots ────────────────────────────────────────
  p_sil_km <- fviz_nbclust(data, kmeans, method = "silhouette",
                            k.max = k_max) +
    labs(title = paste("Avg Silhouette Width — K-Means |", label))

  p_sil_fkm <- fviz_nbclust(data, FKM2, method = "silhouette",
                             k.max = k_max) +
    labs(title = paste("Avg Silhouette Width — Fuzzy K-Means |", label))

  # Print to document
  print(p_gap_km);  print(p_gap_fkm)
  print(p_sil_km);  print(p_sil_fkm)

  # Save as PNG
  save_plot_gg(p_gap_km,  paste0("gap_kmeans_",      clean_label))
  save_plot_gg(p_gap_fkm, paste0("gap_fkm_",         clean_label))
  save_plot_gg(p_sil_km,  paste0("avgSil_kmeans_",   clean_label))
  save_plot_gg(p_sil_fkm, paste0("avgSil_fkm_",      clean_label))

  list(k_km = k_km, k_fkm = k_fkm, k_pdc = k_pdc, k_hc = k_hc)
}
```

### Clustering Functions

Following patterns from kmeans.r (line 82), Fuzzy.R (lines 12–31), and PAM.R (line 85).

```{r}
run_clustering <- function(data, optimal_ks, seed = 420) {
  set.seed(seed)

  # K-Means with its own optimal K
  km  <- kmeans(data, centers = optimal_ks$k_km,  nstart = 10)

  # Fuzzy K-Means with its own optimal K
  fkm <- FKM(X = data, k = optimal_ks$k_fkm, m = 2)

  # PD Clustering with its own optimal K
  pdc <- PDC(data, k = optimal_ks$k_pdc)

  # Hierarchical Average with its own optimal K
  hc     <- agnes(data, method = "average")
  hc_cut <- cutree(as.hclust(hc), k = optimal_ks$k_hc)

  list(km = km, fkm = fkm, pdc = pdc, hc = hc, hc_cut = hc_cut,
       optimal_ks = optimal_ks,
       labels = list(
         km  = as.integer(km$cluster),
         fkm = as.integer(fkm$clus[, 1]),
         pdc = as.integer(pdc$label),
         hc  = as.integer(hc_cut)
       ))
}
```

### Internal Validation Function

Silhouette and CH from lecture (kmeans.r lines 85, 16). Dunn index via `clValid` (clustervalidation.R line 54). Agglomerative Coefficient from `agnes` object.

```{r}
internal_validation <- function(data, clust_out, label) {

  dist_mat <- dist(data, method = "euclidean")
  ks       <- clust_out$optimal_ks

  # ── Silhouette ────────────────────────────────────────────────────────────
  sil_km  <- silhouette(clust_out$km$cluster,    dist_mat)
  sil_fkm <- silhouette(clust_out$fkm$clus[, 1], dist_mat)
  sil_pdc <- silhouette(clust_out$pdc$label,      dist_mat)
  sil_hc  <- silhouette(clust_out$hc_cut,         dist_mat)

  # ── CH Index ──────────────────────────────────────────────────────────────
  hc_centers <- do.call(rbind, lapply(unique(clust_out$hc_cut), function(g)
    colMeans(data[clust_out$hc_cut == g, , drop = FALSE])))

  ch_km  <- ratio.strength(data, ks$k_km,
               list(centers = clust_out$km$centers,
                    cluster  = clust_out$km$cluster))
  ch_fkm <- ratio.strength(data, ks$k_fkm,
               list(centers = clust_out$fkm$H,
                    cluster  = clust_out$fkm$clus[, 1]))
  ch_pdc <- ratio.strength(data, ks$k_pdc,
               list(centers = clust_out$pdc$center,
                    cluster  = clust_out$pdc$label))
  ch_hc  <- ratio.strength(data, ks$k_hc,
               list(centers = hc_centers,
                    cluster  = clust_out$hc_cut))

  # ── Dunn Index ────────────────────────────────────────────────────────────
  dunn_km  <- dunn(dist_mat, clust_out$km$cluster)
  dunn_fkm <- dunn(dist_mat, clust_out$fkm$clus[, 1])
  dunn_pdc <- dunn(dist_mat, clust_out$pdc$label)
  dunn_hc  <- dunn(dist_mat, clust_out$hc_cut)

  # ── Agglomerative Coefficient ─────────────────────────────────────────────
  ac <- clust_out$hc$ac

  # ── Results Table — K column added ───────────────────────────────────────
  results <- tibble(
    Method     = c("K-Means", "Fuzzy K-Means", "PD Clustering", "Hierarchical Avg"),
    K          = c(ks$k_km, ks$k_fkm, ks$k_pdc, ks$k_hc),
    Silhouette = round(c(mean(sil_km[, 3]),  mean(sil_fkm[, 3]),
                         mean(sil_pdc[, 3]), mean(sil_hc[, 3])), 4),
    CH_Index   = round(c(ch_km, ch_fkm, ch_pdc, ch_hc), 4),
    Dunn_Index = round(c(dunn_km, dunn_fkm, dunn_pdc, dunn_hc), 4),
    Agg_Coef   = c(NA, NA, NA, round(ac, 4))
  )

# ── Silhouette Plots ──────────────────────────────────────────────────────
  clean_label <- gsub("[^A-Za-z0-9]", "_", label)

  sil_list <- list(
    list(sil = sil_km,  col = "steelblue", method = "KMeans",
         k = ks$k_km),
    list(sil = sil_fkm, col = "tomato",    method = "FuzzyKMeans",
         k = ks$k_fkm),
    list(sil = sil_pdc, col = "darkgreen", method = "PDClustering",
         k = ks$k_pdc),
    list(sil = sil_hc,  col = "purple",    method = "HierarchicalAvg",
         k = ks$k_hc)
  )

  for (sp in sil_list) {
    title_text <- paste0(gsub("([A-Z])", " \\1", sp$method),
                         " Silhouette | k=", sp$k, " | ", label)
    filename   <- paste0("sil_", sp$method, "_k", sp$k, "_", clean_label)

    # Save PNG
    save_plot_base(
      plot(sp$sil, main = title_text, col = sp$col),
      filename = filename
    )

    # Also render in document
    par(mar = c(5, 4, 4, 10))
    plot(sp$sil, main = title_text, col = sp$col)
  }

  par(mar = c(5, 4, 4, 2))
   # ── Return results and silhouette objects ─────────────────────────────────
  invisible(list(
    results = results,
    sils    = list(km  = sil_km,
                   fkm = sil_fkm,
                   pdc = sil_pdc,
                   hc  = sil_hc),
    ks      = ks
  ))
}
```

### External Validation: ARI

ARI used as external measure comparing clustering methods against each other, following lecture PAM.R (line 118) using `MixGHD::ARI()`.

```{r}
# ARI pairwise comparison across methods — PAM.R line 118
ari_comparison <- function(clust_out, label) {

  km_cl  <- clust_out$km$cluster
  fkm_cl <- clust_out$fkm$clus[, 1]
  pdc_cl <- clust_out$pdc$label
  hc_cl  <- clust_out$hc_cut

  ari_table <- tibble(
    Comparison              = c("K-Means vs Fuzzy K-Means",
                                "K-Means vs PD Clustering",
                                "K-Means vs Hierarchical",
                                "Fuzzy K-Means vs PD Clustering",
                                "Fuzzy K-Means vs Hierarchical",
                                "PD Clustering vs Hierarchical"),
    ARI = round(c(
      ARI(km_cl,  fkm_cl),
      ARI(km_cl,  pdc_cl),
      ARI(km_cl,  hc_cl),
      ARI(fkm_cl, pdc_cl),
      ARI(fkm_cl, hc_cl),
      ARI(pdc_cl, hc_cl)
    ), 4)
  )

  cat("\n=== ARI Table —", label, "===\n")

  # print(
  #   kable(ari_table, longtable = FALSE,
  #         caption  = paste("Pairwise ARI Between Clustering Methods —", label),
  #         booktabs = TRUE) |>
  #     kable_styling(latex_options = "HOLD_position")
  # )

  invisible(ari_table)
}
```

### Master Loop: Run Everything Per Dataset

```{r}
all_internal <- list()
all_ari      <- list()
all_sils     <- list()
all_clust_labels <- list()

for (d in dataset_list) {
  cat("\n\nProcessing:", d$label, "\n")

  data       <- prep_cluster_data(d$df)
  optimal_ks <- get_optimal_k(data, label = d$label, k_max = 8)

  cat("  Optimal K — KM:", optimal_ks$k_km,
      "| FKM:", optimal_ks$k_fkm,
      "| PDC:", optimal_ks$k_pdc,
      "| HC:",  optimal_ks$k_hc, "\n")

  clust_out    <- run_clustering(data, optimal_ks)
  
  # Store cluster labels for cross-dataset ARI
  all_clust_labels[[d$label]] <- clust_out$labels
  
  int_res_full <- internal_validation(data, clust_out, d$label)
  ari_res      <- ari_comparison(clust_out, d$label)

  # Clean label for use in filenames — remove spaces and special chars
  clean_label <- gsub("[^A-Za-z0-9]", "_", d$label)

# ── Internal validation table ─────────────────────────────────────────────
  int_kable <- kable(int_res_full$results, longtable = FALSE,
                     caption  = paste("Internal Validation Measures —", d$label),
                     booktabs = TRUE) |>
    kable_styling(latex_options = c("HOLD_position", "scale_down"),
                  font_size = 9)

  int_kable |> print()
  save_table_png(data         = int_res_full$results,
                 filename     = paste0("internal_validation_", clean_label),
                 caption_text = paste("Internal Validation Measures —", d$label))

  # ── ARI table ─────────────────────────────────────────────────────────────
  ari_kable <- kable(ari_res, longtable = FALSE,
                     caption  = paste("Pairwise ARI —", d$label),
                     booktabs = TRUE) |>
    kable_styling(latex_options = "HOLD_position")

  ari_kable |> print()
  save_table_png(data         = ari_res,
                 filename     = paste0("ari_", clean_label),
                 caption_text = paste("Pairwise ARI —", d$label))

  all_internal[[d$label]] <- int_res_full$results |> mutate(Dataset = d$label)
  all_ari[[d$label]]      <- ari_res              |> mutate(Dataset = d$label)
  all_sils[[d$label]]     <- list(sils  = int_res_full$sils,
                                  ks    = int_res_full$ks,
                                  label = d$label)
}
```

```{r}
#| fig-height: 5
#| fig-width: 9
#| fig-keep: all
#| out-width: "95%"

for (s in all_sils) {
  ks    <- s$ks
  label <- s$label

  par(mar = c(5, 4, 4, 10))
  plot(s$sils$km,
       main = paste0("K-Means Silhouette | k=", ks$k_km, " | ", label),
       col  = "steelblue")

  par(mar = c(5, 4, 4, 10))
  plot(s$sils$fkm,
       main = paste0("Fuzzy K-Means Silhouette | k=", ks$k_fkm, " | ", label),
       col  = "tomato")

  par(mar = c(5, 4, 4, 10))
  plot(s$sils$pdc,
       main = paste0("PD Clustering Silhouette | k=", ks$k_pdc, " | ", label),
       col  = "darkgreen")

  par(mar = c(5, 4, 4, 10))
  plot(s$sils$hc,
       main = paste0("Hierarchical Avg Silhouette | k=", ks$k_hc, " | ", label),
       col  = "purple")
}

par(mar = c(5, 4, 4, 2))  # reset margins
```

### Cross-Dataset Summary Tables

\newpage

```{r}
#| results: asis

# ── Internal validation summary ───────────────────────────────────────────────
int_summary_data <- bind_rows(all_internal) |> relocate(Dataset)

int_summary_data |>
  kable(longtable = FALSE,
        caption   = "Internal Validation Summary Across All Datasets",
        booktabs  = TRUE,
        digits    = 4) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down"),
                font_size = 9) |>
  # Group rows by dataset label instead of collapse_rows
  group_rows(index = setNames(
    rep(4, length(unique(int_summary_data$Dataset))),
    unique(int_summary_data$Dataset)
  )) |>
  print()

save_table_png(
  data         = int_summary_data,
  filename     = "internal_validation_summary_all",
  caption_text = "Internal Validation Summary Across All Datasets",
  group_col    = "Dataset",
  group_size   = 4          # 4 methods per dataset
)

# ── ARI summary ───────────────────────────────────────────────────────────────
ari_summary_data <- bind_rows(all_ari) |> relocate(Dataset)

ari_summary_data |>
  kable(longtable = FALSE,
        caption   = "Pairwise ARI Summary Across All Datasets",
        booktabs  = TRUE,
        digits    = 4) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down"),
                font_size = 9) |>
  # Group rows by dataset label instead of collapse_rows
  group_rows(index = setNames(
    rep(6, length(unique(ari_summary_data$Dataset))),
    unique(ari_summary_data$Dataset)
  )) |>
  print()

save_table_png(
  data         = ari_summary_data,
  filename     = "ari_summary_all",
  caption_text = "Pairwise ARI Summary Across All Datasets",
  group_col    = "Dataset",
  group_size   = 6          # 6 comparisons per dataset
)
```

```{r}
#| results: asis

# ── Cross-dataset ARI within each method — long format ───────────────────────
cross_ari_data <- cross_dataset_ari(all_clust_labels) |>
  mutate(Comparison = paste0(Dataset_1, " vs ", Dataset_2)) |>
  dplyr::select(Method, Comparison, ARI)

n_pairs   <- length(unique(cross_ari_data$Comparison))
n_methods <- length(unique(cross_ari_data$Method))

# ── Print to PDF grouped by method ───────────────────────────────────────────
cross_ari_data |>
  kable(longtable = FALSE,
        caption   = "Cross-Dataset ARI Within Each Clustering Method",
        booktabs  = TRUE,
        digits    = 4,
        col.names = c("Method", "Dataset Comparison", "ARI")) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down"),
                font_size     = 9) |>
  group_rows(index = setNames(
    rep(n_pairs, n_methods),
    unique(cross_ari_data$Method)
  )) |>
  print()

# ── Save one PNG per method to keep tables readable ──────────────────────────
for (m in unique(cross_ari_data$Method)) {

  method_data  <- cross_ari_data |>
    filter(Method == m) |>
    dplyr::select(Comparison, ARI)

  clean_method <- gsub("[^A-Za-z0-9]", "_", m)

  save_table_png(
    data         = method_data,
    filename     = paste0("cross_dataset_ari_", clean_method),
    caption_text = paste("Cross-Dataset ARI —", m)
  )
}
```
